Do motif roles provide unique information about species risk of extinction?

### 2021-06-28 manuscript/motifs_vs_degree.tex

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad of species in food webs depend for their survival \citep{}. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{} and secondary production \citep{}, soil respiration and carbon cycling \citep{} and consumer diversity \citep{scherber2010bottom,Baiser2013}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community promotes stability across all different levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers, e.g. herbivores, predators, and parasitoids \citep{scherber2010bottom}.
     %This has been shown in both theoretical work~\citep{} and empirical experiments ~\citep{} and results from the vertical propagation of disturbances through the network of interacting species.

@@

    %\subsection{6. only 4 (stable) motifs can appear in Bayesian networks, so how do we expect them to affect stability in a bottom-up context?}
    Of the 13 three-species motifs that can occur in empirical food webs there are four motifs that are over-represented in empirical networks \citep{Stouffer2007} and contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}: the three-species chain, omnivory, apparent competition and direct competition motifs.
    These four motifs are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. Because of this restriction, we only consider these four particularly important motifs.  
    % only four do not include cycles (e.g., A eats B and B eats A or A eats B, B eats C, and C eats A). 
    %As cycles must be removed from networks before calculating extinction probabilities in the Bayesian network framework, this study focuses on these four Bayesian-compatible motifs:

@@

            By fitting linear regressions of persistence against motif proportions for the species for each network separately, we can observe how the relationships between motifs and persistence change with the level of disturbance and global network structure. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here. 
            Networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). 
            However, there was an interaction disturbance and connectance for the direct competition and omnivory motifs.

            At low and medium connectances there is little change in the proportion of positive slopes (\textgreater80\%) for the different levels of disturbance. 
            At high connectances, however, the proportion of positive slopes is lower overall and decreases with increasing disturbance. 
            This means that at low levels of connectance, an increased proportion of the direct competition motif increases persistence in most networks, while the opposite is true with high connectance. 
            The trend is reversed for the omnivory motif.

@@

            \begin{figure}
                \centering
                \includegraphics[width=\textwidth]{figures/roles_vs_TL.eps}
                \caption{As with motifs, persistence varied with simpler measures of species roles in networks. Moreover, motif frequencies correlated with in-degree and trophic level. \textbf{A-B)} Persistence increased with increasing in-degree for low levels of disturbance to basal resources but decreased with increasing degree for high levels of disturbance to basal resources.
                Persistence declined with inceasing STL for all levels of disturbance to basal resources.
                \textbf{C-D)} Counts of all motifs increased with increasing in-degree and decreased with increasing STL. The proportion of omnivory increased with increasing degree while all other proportions decreased. The proportions of omnivory and direct competition decreased with increasing STL while the proportions of apparent competition and three-species chains increased.}
                \label{fig:motifs_vs_TL_and_deg}
            \end{figure}

@@

        A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table SX, \emph{Appendix SH}).
        In the omnivory motif, persistence increased with the frequency of the bottom position (two predators, no prey within the motif) across all levels of disturbance but most steeply at high levels of disturbance (Fig.~\ref{figs:persistence_position_motifs}).
        Peristence increased slightly with the frequency of the middle position (one prey, one predator) at low levels of disturbance but showed little relationship at higher levels of disturbance.
        Persistence increased with the frequency of the top position (two prey, no predators within the motif) at low disturbance but decreased at higher levels of disturbance.

@@

Somewhere early in discusssion (needs tidying up):

Apparent comp and direct comp have no indirect effects in BN, so we should be able to explain them with simpler properties.
AC is strongly negatively correlated with in-degree. Higher in degree is associated with both high persistence at low disturbance (Fig. 7A) and low participation in AC (Fig. 7C). Low participation in AC is in turn associated with high persistence at low disturbance (Fig. 4). Note that trends for degree and AC reverse at high disturbance in the same way!
DC is not strongly correlated with degree but does decrease strongly at high trophic levels. High trophic level and low DC are both associated with low persistence at all levels of disturbance, making it likely that the effects we observe for DC are trophic level in disguise.

Three-species chain and omnivory do have indirect effects in a BN setting, and they can't be fully explained by degree or trophic level (since the result doesn't follow the result for degree). We can see this because despite the strong correlations with degree and trophic level these motifs don't show the same patterns with persistence. We therefore focus on them here.

@@

%In all cases, the effect of increasing disturbance to basal resources had a much larger effect on consumer persistences than did network structure. This is a consequence of the bottom-up calculations done to receive the final persistence. The importance of the basal level for network robustness has been demonstrated in numerous studies \citep{scherber2010bottom}, as well as the sensitivity of high trophic levels (REF).

@@

As well as the mean persistence of all consumers, we also considered the persistence of each consumer in a network.
While the persistence of a particular species is likely to be affected by global-scale (size and connectance) and meso-scale (motif profiles) network structure, it will also depend on species-specific characteristics.
In particular, a species' motif participation was significantly related to its probability of persistence.
Specifically, a high proportion of the direct competition or omnivory motifs increase the probability of persistence, while a high proportion of the three-species chain motif decreases probability of persistence.
%These trends held at any level of disturbance we considered, although at high levels of disturbance the positive effect of omnivory was reduced and the negative effect of three-species chains was increased.
%The effect of the apparent competition motif, however, varied with disturbance.
%A high proportion of the apparent competition motif reduces persistence at low levels of disturbance but increases persistence at high levels of disturbance.

% Two notable results: 1) omnivory bad at network level, good at species level, 2) apparent competition depends on disturbance
Combining these results with our network-level findings, we see that while being in a network with high frequencies of omnivory decreases persistence, participating in many of these omnivory motifs increases persistence at the species level.
At the network level, a higher frequency of the omnivory motif leads to more pathways by which disturbances to basal resources can propagate and cause extinctions of consumers.
At a species level, if extinctions of basal resources are rare, these multiple pathways will tend to protect consumers participating in many omnivory motifs by providing alternative channels of energy.
If extinctions of basal resources are common, however, these multiple pathways cause a negative impact as they will transmit more disturbances to the consumer.
To understand the species-level effect, we consider the relationships between stability and each position within the motif.
Appearing frequently in the top or middle positions of the omnivory motif is associated with higher probabilities of persistence at low levels of disturbance, while appearing frequently in the bottom position is associated with higher probabilities of persistence at high levels of disturbance.
This is consistent with the logic outlined for the network-level effect: when extinction of basal species is rare, consumers at the top or middle of omnivory motifs benefit from multiple energy channels but when extinction is common these multiple channels mean that the consumer is likely to be affected by multiple disturbances.
The bottom position shows the reverse effect: at low levels of disturbance, this position is nearly neutral. 
At high levels of disturbance, however, being low down in the motif (and obtaining energy by appearing as a predator in other motifs) means that the species is less likely to be affected by multiple lower-level extinctions.
The offsetting of effects of different positions within the omivory motif may explain why the motif as a whole is associated with greater probabilities of persistence across all levels of disturbance we considered.

%The different positions in the three-species chain motif - connect to network stability? 3-sp chain always good in fig 3 - when a full food web has a high proportion of this motif. In fig 4, a species with a high proportion of the 3-sp chain always negative for persistence, most so with increasing disturbance. In fig 8, mainly the top position that reduces persistence, more so with high disturbance. Bottom and middle position largely unchanged.

% in app comp - bottom species drive the trend visible in fig 4? the top species shows opposite patterns anyway.

% % Might be able to add some of this paragraph to the one above or below (especially the refs).
% The divergent effect on persistence of either having a high proportion of the three-species chain or omnivory motif can, within our Bayesian Network framework, be explained by the structure of the food web.
% When participating in a food chain motif---three-species chain or omnivory---the top species of this structure will have a higher susceptibility since disturbances amplify through the network. In our results, this is in particular evident for the three-species chain motif at higher levels of disturbance, but this also applies to species with a high proportion of the omnivory motif. While the relationship between omnivory and persistence is strongly positive at low bottom-level disturbance, the trend is reduced at high levels of disturbance. 
% This is because species participating to a high extent in omnivory motifs share the sensitivity of species participating to a high extent in three-species chain motif, and are similarly negatively affected by bottom-level disturbances. However, the top species in an omnivory motif also have the benefit of an additional prey species, located further down in the food web. This prey item can assist as a buffer, decreasing the negative effect of disturbance on the consumer species. Clearly, this additional prey is of importance, since it is more beneficial with a higher proportion of the omnivory motif for an individual species than a high proportion of the three-species chain motif. 
% Thus, the observed benefit of a high proportion of the omnivory motif can be compared to omnivores in a network feeding adaptively. Studies have demonstrated that adaptive omnivory can increase the permanence of three-species chains \citep{Fagan1997, KrivanDiehl2005, AbramsFung2010}.
% Although species in our model cannot swap prey dependent on bottom-level disturbances or lower resource densities, a species with a high proportion of omnivory will in a sense spread its risk between prey items at different trophic levels.
% Additionally, the effect of the omnivory motif on persistence was closely connected to connectance. In highly connected networks, a high proportion of the omnivory motif strongly contributed to higher persistence. This was not the case in less connected networks. 
% In a highly connected network, a disturbance could potentially spread to a higher extent than in a less connected network. 
% Therefore, in a highly connected web where disturbances have more ways to move, the lower level prey item in the omnivory motif may be of even higher importance assisting as a buffer, decreasing the negative effect of disturbance for the consumer species. In a less connected network, this additional prey species is of less importance.

@@

Our result contrasts with \citep{Cirtwill2021_inprep}, where a high proportion of the omnivory motif resulted in a shorter time to extinction, while the other three stable motifs were associated with longer times to extinction. However, the study by Cirtwill and Wootton used a different simulation approach, with body-mass structured vertebrate food webs and a dynamic simulation model accounting for top-down effects. 
    %In previous studies of the effect of omnivory motifs, the strength of the omnivorous link can potentially be important. If this link is strong, it is shown to be destabilizing (REF), while stabilizing if weaker (REF). In our Bayesian Network setting, this link is indistinguishable from any other link in the food web which partly may explain why omnivory results in increased persistence of consumer species.

@@

\bibliographystyle{ecollett} 
\bibliography{anna_bib} % Abbreviate journal titles.

>>>>

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad of species in food webs depend for their survival \citep{}. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{} and secondary production \citep{}, soil respiration and carbon cycling \citep{} and consumer diversity \citep{scherber2010bottom, Baiser2016}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community promotes stability across all different levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers, e.g. herbivores, predators, and parasitoids \citep{scherber2010bottom}.
     %This has been shown in both theoretical work~\citep{} and empirical experiments ~\citep{} and results from the vertical propagation of disturbances through the network of interacting species.

@@

    %\subsection{6. only 4 (stable) motifs can appear in Bayesian networks, so how do we expect them to affect stability in a bottom-up context?}
    Of the 13 three-species motifs that can occur in empirical food webs there are four motifs that are over-represented in empirical networks \citep{Stouffer2007} and contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}: the three-species chain, omnivory, apparent competition and direct competition motifs.
    These four motifs are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. Because of this restriction, we only consider these four particularly important motifs.

@@

            By fitting linear regressions of persistence against motif proportions for the species for each network separately, we can observe how the relationships between motifs and persistence change with the level of disturbance and global network structure. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here. 
            Networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). 
            However, there was an interaction between disturbance and connectance for the direct competition and omnivory motifs.

            At low and medium connectances with a high proportion of the direct competition motif, there is little change in the proportion of positive slopes (\textgreater80\%) for the different levels of disturbance. 
            At high connectances, however, the proportion of positive slopes is lower overall and decreases with increasing disturbance. 
            This means that at low levels of connectance, an increased proportion of the direct competition motif increases persistence in most networks, while the opposite is true with high connectance. 
            The trend is reversed for the omnivory motif.

@@

            \begin{figure}
                \centering
                \includegraphics[width=\textwidth]{figures/roles_vs_TL.eps}
                \caption{As with motifs, persistence varied with simpler measures of species roles in networks. Moreover, motif frequencies correlated with in-degree and trophic level. \textbf{A-B)} Persistence increased with increasing in-degree for low levels of disturbance to basal resources but decreased with increasing degree for high levels of disturbance to basal resources.
                Persistence declined with inceasing STL for all levels of disturbance to basal resources.
                \textbf{C-D)} The proportion of omnivory increased with increasing degree while all other proportions decreased. The proportions of omnivory and direct competition decreased with increasing STL while the proportions of apparent competition and three-species chains increased.}
                \label{fig:motifs_vs_TL_and_deg}
            \end{figure}

@@

        A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table SX, \emph{Appendix SH}).
        In the omnivory motif, persistence increased with the frequency of the bottom position (two predators, no prey within the motif) across all levels of disturbance but most steeply at high levels of disturbance (Fig.~\ref{figs:persistence_position_motifs}).
        Persistence increased slightly with the frequency of the middle position (one prey, one predator) at low levels of disturbance but showed little relationship at higher levels of disturbance.
        Persistence increased with the frequency of the top position (two prey, no predators within the motif) at low disturbance but decreased at higher levels of disturbance.

@@

%Somewhere early in discussion (needs tidying up):
%Apparent comp and direct comp have no indirect effects in BN, so we should be able to explain them with simpler properties.
%AC is strongly negatively correlated with in-degree. Higher in degree is associated with both high persistence at low disturbance (Fig. 7A) and low participation in AC (Fig. 7C). Low participation in AC is in turn associated with high persistence at low disturbance (Fig. 4). Note that trends for degree and AC reverse at high disturbance in the same way!
%DC is not strongly correlated with degree but does decrease strongly at high trophic levels. High trophic level and low DC are both associated with low persistence at all levels of disturbance, making it likely that the effects we observe for DC are trophic level in disguise.
%Three-species chain and omnivory do have indirect effects in a BN setting, and they can't be fully explained by degree or trophic level (since the result doesn't follow the result for degree). We can see this because despite the strong correlations with degree and trophic level these motifs don't show the same patterns with persistence. We therefore focus on them here.

@@

@@

% HERE ENDS MEAN PERSISTENCE AND SPECIES WISE START
As well as the mean persistence of all consumers, we also considered the species-specific persistence of each consumer in a network.
While the persistence of a particular species is likely to be affected by global-scale (size and connectance) and meso-scale (motif profiles) network structure, it also depends on a species' position in the network. 
There are many ways of measuring species' positions, and in some cases multiple measures may capture the same trend.
In our results, motif participation is significantly related to persistence, but the effects of two motifs (\emph{apparent competition} and \emph{direct competition}) might be explained more simply by in-degree and trophic level.
This is especially likely for these two motifs, since Bayesian Network modeling lacks top-down effects and therefore removes the indirect effects from these motifs.

% Getting rid of AC/DC "quickly" here so we can focus on omnivory
Participation in the apparent competition motif is strongly negatively correlated with in-degree---a high proportion of the apparent competition motif implies a low in-degree. 
Both higher in-degree and lower participation in the apparent competition motif are associated with greater persistence at low disturbance but decreased persistence at high disturbance; because the trends are so similar, it is likely that a common mechanism underlies both.
% The effect on persistence is consistent between species with a low in-degree and a high proportion of the apparent competition motif. 
%Low in-degree decreases persistence at low disturbance, and increases persistence at high disturbance, similar to the trend visible for a high participation in the apparent competition motif. 
% Similarly, the effect on species persistence is reversed for a low proportion of the apparent competition motif in which species have a high in-degree.

While the direct competition motif is not correlated with in-degree, it is strongly negatively correlated with trophic level---species with a high proportion of the direct competition motif have a low trophic levels. 
Low trophic levels are always associated with higher persistence because these species are less affected by bottom-up disturbances~\citep{}.
It then follows that species which participate in many direct competition motifs may have higher probabilities of persistence because they are also likely to have low trophic levels.

In contrast to the competition motifs, the three-species chain and omnivory motifs do have indirect effects in a Bayesian Network setting and the effects of these motifs cannot be fully explained by degree or trophic level.
For example, the three-species chain motif is negatively correlated with in-degree and positively correlated with trophic level. 
Despite this correlation, participation in the three-species chain does not interact the same way with disturbance as either simpler measure and so has a unique relationship to species persistence.
This is likely due to indirect effects of the bottom species in the chain on the top species.

The most interesting motif in which to consider indirect effects is omnivory.
While participation in the omnivory motif is positively correlated with in-degree, a high proportion of the omnivory motif is always associated with increased persistence, while the effect of in-degree varies with disturbance. 
Thus, the effect of a high proportion of the three-species chain motif and the omnivory motif does not follow the corresponding effect of in-degree and trophic level.

% Two notable results: 1) omnivory bad at network level, good at species level, 2) apparent competition depends on disturbance
Combining the results from our species wise exploration with our network-level findings, we see that while being in a network with high frequencies of omnivory decreases persistence, participating in many of these omnivory motifs increases persistence at the species level.
At the network level, a higher frequency of the omnivory motif leads to more pathways by which the disturbances to basal resources can propagate and cause extinctions of consumers.
At the species level on the other hand, if extinctions of basal resources are rare, these multiple pathways will tend to protect consumers participating in many omnivory motifs by providing alternative channels of energy.
If extinctions of basal resources are common, however, these multiple pathways cause a negative impact as they will transmit more disturbances to the consumer.

Additionally, the effect of the omnivory motif on persistence was closely connected to connectance. In highly connected networks, a high proportion of the omnivory motif strongly contributed to higher persistence. This was not the case in less connected networks. 
In a highly connected network, a disturbance could potentially spread to a higher extent than in a less connected network. 
Therefore, in a highly connected web where disturbances have more ways to move, the lower level prey item in the omnivory motif may be of even higher importance assisting as a buffer, decreasing the negative effect of disturbance for the consumer species. In a less connected network, this additional prey species is of less importance.

To understand the species-level effect, we consider the relationships between stability and each position within the motif.
Appearing frequently in the top or middle positions of the omnivory motif is associated with higher probabilities of persistence at low levels of disturbance, while appearing frequently in the bottom position is associated with higher probabilities of persistence at high levels of disturbance.
This is consistent with the logic outlined for the network-level effect: when extinction of basal species is rare, consumers at the top or middle of omnivory motifs benefit from multiple energy channels but when extinction is common these multiple channels mean that the consumer is likely to be affected by multiple disturbances. The bottom position shows the reverse effect: at low levels of disturbance, this position is nearly neutral.  At high levels of disturbance, however, being low down in the motif (and obtaining energy by appearing as a predator in other motifs) means that the species is less likely to be affected by multiple lower-level extinctions.
The offsetting of effects of different positions within the omivory motif may explain why the motif as a whole is associated with greater probabilities of persistence across all levels of disturbance we considered.

% % Might be able to add some of this paragraph to the one above or below (especially the refs).
% The divergent effect on persistence of either having a high proportion of the three-species chain or omnivory motif can, within our Bayesian Network framework, be explained by the structure of the food web.
% When participating in a food chain motif---three-species chain or omnivory---the top species of this structure will have a higher susceptibility since disturbances amplify through the network. In our results, this is in particular evident for the three-species chain motif at higher levels of disturbance, but this also applies to species with a high proportion of the omnivory motif. While the relationship between omnivory and persistence is strongly positive at low bottom-level disturbance, the trend is reduced at high levels of disturbance. 
% This is because species participating to a high extent in omnivory motifs share the sensitivity of species participating to a high extent in three-species chain motif, and are similarly negatively affected by bottom-level disturbances. However, the top species in an omnivory motif also have the benefit of an additional prey species, located further down in the food web. This prey item can assist as a buffer, decreasing the negative effect of disturbance on the consumer species. Clearly, this additional prey is of importance, since it is more beneficial with a higher proportion of the omnivory motif for an individual species than a high proportion of the three-species chain motif. 
% Thus, the observed benefit of a high proportion of the omnivory motif can be compared to omnivores in a network feeding adaptively. Studies have demonstrated that adaptive omnivory can increase the permanence of three-species chains \citep{Fagan1997, KrivanDiehl2005, AbramsFung2010}.
% Although species in our model cannot swap prey dependent on bottom-level disturbances or lower resource densities, a species with a high proportion of omnivory will in a sense spread its risk between prey items at different trophic levels.

@@

 \cite{Cirtwill2021_inprep} analysed how species time to extinction varied depending on their network profile. They found that a high proportion of the omnivory motif resulted in a shorter time to extinction, while the other three stable motifs were associated with longer times to extinction. However, the study by Cirtwill and Wootton used a different approach using a dynamic simulation model also including top-down effects. 
    %In previous studies of the effect of omnivory motifs, the strength of the omnivorous link can potentially be important. If this link is strong, it is shown to be destabilizing (REF), while stabilizing if weaker (REF). In our Bayesian Network setting, this link is indistinguishable from any other link in the food web which partly may explain why omnivory results in increased persistence of consumer species.

@@

\bibliographystyle{ecollett} 
\bibliography{anna_bib_new} % Abbreviate journal titles.
